Ecological management of stochastic systems with long transients

### 2020-01-04 manuscript/manuscript.Rmd
add line numbers
repo organization, structure

---
title: "Ecological management of stochastic systems with long transients"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  Long transients may be common in ecological dynamics, but the implications such dynamics have for ecological management have not been fully explored.  Long transient periods can easily be mistaken for stable-state dynamics, but may require dramatically different management policies.  Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: a recently recognized mechanism which can give rise to long transients. I consider three approaches of increasing sophistication: (1) dynamic management under a fixed model, (2) management that accounts for uncertainty over possible models, and (3) adaptive management that actively learns the correct model over the management process. This analysis confirms the prediction that long transients can create considerable uncertainty and give rise to very different optimal management policies, but also illustrates that dynamic management that can either plan for this uncertainty or actively learn to decrease the uncertainty can promote successful management of long transients.
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
preamble: |
  \journal{Theoretical Ecology}
  \usepackage[left]{lineno}

@@

```{r figure8, fig.cap="Learning through Adaptive Management of the forest-beetle system. Bars show the probabilities that each model is correct over the first 10 steps of the simulated management process shown in Fig 7.  Initial probabilities for each model are equal, and learning proceeds over the first 5 timesteps (shown in lighter shades in decreasing probability of the bistable model beign correct).  By the fifth iteration, the manager has opted to eliminate the beetles and further learning is no longer possible in this system."}
model_names <- c("Ghost", "Bistable")
learning$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% floor(seq(1,10, length.out=5))) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = model_names[as.integer(param)]) %>% 
  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1])
```

@@

>>>>

---
title: "Ecological management of stochastic systems with long transients"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  Long transients may be common in ecological dynamics, but the implications such dynamics have for ecological management have not been fully explored.  Long transient periods can easily be mistaken for stable-state dynamics, but may require dramatically different management policies.  Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: a recently recognized mechanism which can give rise to long transients. I consider three approaches of increasing sophistication: (1) dynamic management under a fixed model, (2) management that accounts for uncertainty over possible models, and (3) adaptive management that actively learns the correct model over the management process. This analysis confirms the prediction that long transients can create considerable uncertainty and give rise to very different optimal management policies, but also illustrates that dynamic management that can either plan for this uncertainty or actively learn to decrease the uncertainty can promote successful management of long transients.
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article
preamble: |
  \journal{Theoretical Ecology}
  \usepackage{lineno}
  \linenumbers

@@

```{r figure8, fig.cap="Learning through Adaptive Management of the forest-beetle system. Bars show the probabilities that each model is correct over the first 10 steps of the simulated management process shown in Fig 7.  Initial probabilities for each model are equal, and learning proceeds over the first 5 timesteps (shown in lighter shades in decreasing probability of the bistable model beign correct).  By the fifth iteration, the manager has opted to eliminate the beetles and further learning is no longer possible in this system."}
model_names <- c("Ghost", "Bistable")
learning$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% floor(seq(1,10, length.out=5))) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = model_names[as.integer(param)]) %>% 
  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
  xlab("Model")
```

@@

# Acknowledgements

This work was inspired and influenced many discussions at the 2019 NIMBioS workshop on transient dynamics in ecology, and the 2019 organized oral session on ecological transients.  CB also acknowledges computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms, as well as the support from UC Berkeley and the USDA Hatch project CA-B-INS-0162-H.
